{\steve}: A hierarchical Bayesian model for Supernova Cosmology

### 2019-03-26 papers/methods/paper.tex
Fixing two typos

One large difference between traditional methods and BHM methods is that we treat selection effects by incorporating selection efficiency into our model, rather than relying on simulation-driven data corrections. We describe the probability that the \added{ possible} events we observe are drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are observed and pass cuts.  To make this extra conditional explicit, we can write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}
As our model so far describes components of a basic likelihood $P({\rm data}|\theta)$, and we wish to formulate a function $P(S|{\rm data},\theta)$ that describes the chance of an event being successfully observed, we rearrange the likelihood in terms of those functions and find
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, 
\end{align}
where the denominator represents an integral over all potential data $D$, \added{and $\theta$ represents top-level parameters. In the case that our selection effects are best charactersied by latent variables \textit{instead of} data, we can add them to our formulation and our likelihood becomes
\begin{eqnarray}
\mathcal{L}(\theta; {\rm data})&=& \frac{\int P(S|L,\theta) P({\rm data}|L) P(L|\theta)\, dL}{\iint P(S|L,\theta) P(L|\theta) dL},\label{eq:main}
\end{eqnarray}
where $L$ represents our latent parameters.}
This is derived in Appendix \ref{app:selection1}. \added{ To evaluate the effect of our selection effects, we need to evaluate both the selection effect terms in the numerator and the integral in the denominator. The numerator represents the probability we caught the supernova and it was selected into the cosmology sample. The integral represents our global selection efficiency at a location in parameter space, rather than the probability of our data being selected into our sample.} As $\theta$ represents the vector of all \added{top-level model parameters, and $L$ represents a vector of all latent parameters, this is not a trivial integral.} Techniques to approximate this integral, such as Monte-Carlo integration or high-dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by Hamiltonian Monte-Carlo, and post-fitting importance sampling failed due to high-dimensionality (a brief dismissal of many months of struggle). We therefore simplify the integral and approximate the selection effects from their full expression in all of $\theta$-space, to apparent magnitude and redshift space independently (not dependent on $x_1$ or $c$), such that the denominator of equation \eqref{eq:main}, denoted now $d$ for simplicity, is given as
\begin{equation}
d = \int  \left[ \int P(S|m_B) P(m_B | z, \theta)\, d m_B \right] P(S|z) P(z|\theta)\, dz, \label{eq:w1}
\end{equation}
where $P(m_B | z, \theta)$ can be expressed by translating the underlying $M_B$, $x_1$, and $c$ population to $m_B$ given cosmological parameters. A full derivation of this can be found in Appendix \ref{app:selection2}.

@@

\subsection{Supernova Selection Effects}
\label{app:selection2}
\added{
To turn the generalised equations from the previous sections into selection effects relevant for this model, need to highlight that $\theta$ represents only our top level parameters ($\Omega_m$, $w$, $\alpha$, $\beta$, etc), and that our parametrisation of true underlying values (for example $m_B$) takes the form of latent parameters. We thus continue from equation \eqref{eq:latent} and write out denominator $d$:
\begin{equation}
d = \iint  P(S|D, L, \theta) P(D | L, \theta) P(L | \theta)\, dD\, dL.
\end{equation}
In our formulation, we assume that our selection effects can be sufficiently encapsulated by latent parameters. That is, we simplify $P(S|D,L, \theta) \rightarrow P(S|L)$. This allows us to isolate our integral $\int P(D|L,\theta)$ in the above equation, integrate it to unity, and come to 
\begin{equation}
d = \int  P(S|L) P(L | \theta)\, dL.
\end{equation}	
Here we can now move from the generic label $L$ to something specific to our model. Specifically, we assume that our selection effects can be quantified using the true apparent magnitude $m_B$ and redshift $z$, and so $L \rightarrow \lbrace m_B, z\rbrace$, or formally $P(S|D, L, \theta) = P(S|z) P(S|m_B)$. We do not write out other latent parameters found in the model as they do not impact selection effects --- we would simply find them integrated to unity. Note that this assumption does not capture biases engendered by
Poisson noise fluctuations.  
We advocate that future analyses with
higher statistical precision do use a precisely determined $P(S|D)$.
Writing out the latent variables, our denominator becomes
\begin{equation}
d = \int P(S|z) P(S|m_B)P(z, m_B | \theta)\, d\hat{z} \, dz \, dm_B.  \label{eq:sel} 
\end{equation}
We can express $P(z, m_B | \theta)$ as  $P(m_B | z, \theta) P(z | \theta)$, where the first term requires us to calculate the magnitude distribution of our underlying population at a given redshift, and the second term is dependent on survey geometry and supernovae rates. We can thus state}
\begin{equation}
d = \int \left[ \int P(S|m_B) P(m_B | z, \theta)\, dm_B \right] P(S|z)P(z|\theta)\, dz.
\end{equation}
By assuming that the distribution $P(S|z)P(z|\theta)$ is well sampled by the observed supernova redshifts, we can approximate the integral over redshift by evaluating
\begin{equation}
\int P(S|m_B) P(m_B | z, \theta)\, dm_B \label{eq:selint}
\end{equation}
for each supernova in the dataset -- i.e. Monte Carlo integration with assumed perfect importance sampling.

>>>>

One large difference between traditional methods and BHM methods is that we treat selection effects by incorporating selection efficiency into our model, rather than relying on simulation-driven data corrections. We describe the probability that the \added{ possible} events we observe are drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are observed and pass cuts.  To make this extra conditional explicit, we can write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}
As our model so far describes components of a basic likelihood $P({\rm data}|\theta)$, and we wish to formulate a function $P(S|{\rm data},\theta)$ that describes the chance of an event being successfully observed, we rearrange the likelihood in terms of those functions and find
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, 
\end{align}
where the denominator represents an integral over all potential data $D$, \added{and $\theta$ represents top-level parameters. In the case that our selection effects are best characterised by latent variables \textit{instead of} data, we can add them to our formulation and our likelihood becomes
\begin{eqnarray}
\mathcal{L}(\theta; {\rm data})&=& \frac{\int P(S|L,\theta) P({\rm data}|L) P(L|\theta)\, dL}{\iint P(S|L,\theta) P(L|\theta) dL},\label{eq:main}
\end{eqnarray}
where $L$ represents our latent parameters.}
This is derived in Appendix \ref{app:selection1}. \added{ To evaluate the effect of our selection effects, we need to evaluate both the selection effect terms in the numerator and the integral in the denominator. The numerator represents the probability we caught the supernova and it was selected into the cosmology sample. The integral represents our global selection efficiency at a location in parameter space, rather than the probability of our data being selected into our sample.} As $\theta$ represents the vector of all \added{top-level model parameters, and $L$ represents a vector of all latent parameters, this is not a trivial integral.} Techniques to approximate this integral, such as Monte-Carlo integration or high-dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by Hamiltonian Monte-Carlo, and post-fitting importance sampling failed due to high-dimensionality (a brief dismissal of many months of struggle). We therefore simplify the integral and approximate the selection effects from their full expression in all of $\theta$-space, to apparent magnitude and redshift space independently (not dependent on $x_1$ or $c$), such that the denominator of equation \eqref{eq:main}, denoted now $d$ for simplicity, is given as
\begin{equation}
d = \int  \left[ \int P(S|m_B) P(m_B | z, \theta)\, d m_B \right] P(S|z) P(z|\theta)\, dz, \label{eq:w1}
\end{equation}
where $P(m_B | z, \theta)$ can be expressed by translating the underlying $M_B$, $x_1$, and $c$ population to $m_B$ given cosmological parameters. A full derivation of this can be found in Appendix \ref{app:selection2}.

@@

\subsection{Supernova Selection Effects}
\label{app:selection2}
\added{
To turn the generalised equations from the previous sections into selection effects relevant for this model, need to highlight that $\theta$ represents only our top level parameters ($\Omega_m$, $w$, $\alpha$, $\beta$, etc), and that our parametrisation of true underlying values (for example $m_B$) takes the form of latent parameters. We thus continue from equation \eqref{eq:latent} and write out denominator $d$:
\begin{equation}
d = \iint  P(S|D, L, \theta) P(D | L, \theta) P(L | \theta)\, dD\, dL.
\end{equation}
In our formulation, we assume that our selection effects can be sufficiently encapsulated by latent parameters. That is, we simplify $P(S|D,L, \theta) \rightarrow P(S|L)$. This allows us to isolate our integral $\int P(D|L,\theta)$ in the above equation, integrate it to unity, and come to 
\begin{equation}
d = \int  P(S|L) P(L | \theta)\, dL.
\end{equation}	
Here we can now move from the generic label $L$ to something specific to our model. Specifically, we assume that our selection effects can be quantified using the true apparent magnitude $m_B$ and redshift $z$, and so $L \rightarrow \lbrace m_B, z\rbrace$, or formally $P(S|D, L, \theta) = P(S|z) P(S|m_B)$. We do not write out other latent parameters found in the model as they do not impact selection effects --- we would simply find them integrated to unity. Note that this assumption does not capture biases engendered by
Poisson noise fluctuations.  
We advocate that future analyses with
higher statistical precision do use a precisely determined $P(S|D)$.
Writing out the latent variables, our denominator becomes
\begin{equation}
d = \int P(S|z) P(S|m_B)P(z, m_B | \theta)\, dz \, dm_B.  \label{eq:sel} 
\end{equation}
We can express $P(z, m_B | \theta)$ as  $P(m_B | z, \theta) P(z | \theta)$, where the first term requires us to calculate the magnitude distribution of our underlying population at a given redshift, and the second term is dependent on survey geometry and supernovae rates. We can thus state}
\begin{equation}
d = \int \left[ \int P(S|m_B) P(m_B | z, \theta)\, dm_B \right] P(S|z)P(z|\theta)\, dz.
\end{equation}
By assuming that the distribution $P(S|z)P(z|\theta)$ is well sampled by the observed supernova redshifts, we can approximate the integral over redshift by evaluating
\begin{equation}
\int P(S|m_B) P(m_B | z, \theta)\, dm_B \label{eq:selint}
\end{equation}
for each supernova in the dataset -- i.e. Monte Carlo integration with assumed perfect importance sampling.
